---
title: "LoL Data Analysis"
author: "Matthew Matsuo"
date: "4/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introductory Analysis to My LoL Data

```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
setwd("C:/Users/Matt Matsuo/Google Drive/Coding/LoL")
library(dplyr)
library(tidyr)
library(ggplot2)


#Import Data from 2018, 2019, and 2020
dat18 <- readxl::read_xlsx("lol18.xlsx")
colnames(dat18)[22] <- "Win1"
colnames(dat18)[23] <- "Loss1"

dat19 <- readxl::read_xlsx("lol19.xlsx")
colnames(dat19)[24] <- "Win1"
colnames(dat19)[25] <- "Loss1"

dat20 <- readxl::read_xlsx("lol20.xlsx")
dat20 <- dat20[1:58,]
colnames(dat20)[25] <- "Win1"
colnames(dat20)[26] <- "Loss1"



#Consolidate Partner names in each dataset
for(i in 1:nrow(dat18)){
  if(dat18$Partner[i] == "clap them cheeks"){
    dat18$Partner[i] = "Vik"
  }
  if(dat18$Partner[i] == "Monoploxy"){
    dat18$Partner[i] = "Vik"
  }
  if(dat18$Partner[i] == "dripking ruth"){
    dat18$Partner[i] = "Vik"
  }
  if(dat18$Partner[i] == "na"){
    dat18$Partner[i] = NA
  }
}

for(j in 1:nrow(dat19)){
  if(dat19$Partner[j] == "ruth" && !is.na(dat19$Partner[j])){
    dat19$Partner[j] = "Vik"
  }
  if(dat19$Partner[j] == "Feline" && !is.na(dat19$Partner[j])){
    dat19$Partner[j] = "Vik"
  }
}
```


## Solo vs Duo Queue
Let's first take a look at how my winrates differ between playing Duo with a friend versus Solo alone.

```{r}
#Display winrates by Partner and over all 3 years
#2018
two <- dat18 %>%
  group_by(Partner) %>%
  count("Wins" = sum(Win1), "Losses" = sum(Loss1))
t18 <- two %>%
  mutate("WinPct" = Wins/(Wins + Losses), 
         "Year" = 2018)

#2019
two2 <- dat19 %>%
  group_by(Partner) %>%
  count("Wins" = sum(Win1), "Losses" = sum(Loss1))
t19 <- two2 %>%
  mutate("WinPct" = Wins/(Wins + Losses),
         "Year" = 2019)

#2020
two3 <- dat20 %>%
  group_by(Partner) %>%
  count("Wins" = sum(Win1), "Losses" = sum(Loss1))
t20 <- two3 %>%
  mutate("WinPct" = Wins/(Wins + Losses),
         "Year" = 2020)

t18
t19
t20
```

The three previous tables show my winrates from 2018, 2019, and 2020 respectively. In 2018, I had an incredible winrate with Vik and also played well Solo (Partner == NA). 2019 displayed similar winrates, though, my winrate with Yondai went down dramatically. So far in 2020, my winrate with Vik has flipped and is very low compared to previous years. My Solo rate is still high though.  
<br>  
<br>    


```{r}
#Combined
dat <- rbind(t18,t19,t20)
dat <- tibble(dat)
x <- dat %>%
  group_by(Partner) %>%
  count("Wins" = sum(Wins), "Losses" = sum(Losses))
x <- x %>%
  mutate("WinPct" = Wins / (Wins + Losses))
x
```

Overall, my winrate in SoloQ is higher than the average of 50%. My winrate with Vik and Chanteon are also greater than 50% while it is worse with Setsuji and Yondai.
<br>
<br>

```{r}
#Plot Winrates by Partner
ddd <- data.frame(dat)

ggplot(data=ddd, aes(x=Partner, y=WinPct, fill=factor(Year))) +
  geom_bar(stat="identity", position=position_dodge()) +
  geom_text(aes(label=n), vjust=-.5, color="black", size=3.5, position = position_dodge(0.9))+
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0:1), breaks = seq(0,1,.1)) +
  geom_hline(yintercept = .5, linetype = "dashed", color = "red") +
  theme_bw() +
  ggtitle("Winrates by Duo Parter") +
  theme(plot.title = element_text(hjust = 0.5))
```

My winrates with Chanteon and Solo (NA) are the only ones that are at or above 50% for all three years.
<br>
<br>

## Winrates by Month
Now, let's investigate if the time of the year has any impact on my winrates.

```{r}
#2018
bymonth18 <- dat18 %>% 
  group_by(Month) %>% 
  summarise(KDA=sum(Kills + Assists)/sum(Deaths), Wins=sum(Win1),WinPct=sum(Win1)/n(),n=n()) %>%
  mutate(Year = 2018) %>% 
  arrange(desc(WinPct)) 

#2019
bymonth19 <- dat19 %>% 
  group_by(Month) %>% 
  summarise(KDA=sum(Kills + Assists)/sum(Deaths), Wins=sum(Win1),WinPct=sum(Win1)/n(),n=n()) %>%
  mutate(Year = 2019) %>% 
  arrange(desc(WinPct)) 

#2020
bymonth20 <- dat20 %>% 
  group_by(Month) %>% 
  summarise(KDA=sum(Kills + Assists)/sum(Deaths), Wins=sum(Win1),WinPct=sum(Win1)/n(),n=n()) %>% 
  mutate(Year = 2020) %>% 
  arrange(desc(WinPct))
bymonth18
bymonth19
bymonth20
```

<br>
<br>

```{r}
#Combined
dat_month <- rbind(bymonth18, bymonth19, bymonth20)
dat_month <- tibble(dat_month)
x2 <- dat_month %>%
  mutate("Losses" = n - Wins)
x2 <- x2 %>%
  group_by(Month) %>%
  summarise(Wins=sum(Wins), Losses=sum(Losses))
x2 <- x2 %>%
  mutate("WinPct" = Wins / (Wins + Losses))
x2

#Plot Winrates by Month
ddd2 <- data.frame(dat_month)
ddd2$Month <- factor(ddd2$Month,levels = c("January", "February", "March", "April","May","June","July","August","September","October","November"))
overallwinrate <- sum(ddd$Wins)/sum(ddd$n)

ggplot(data=ddd2, aes(x=Month, y=WinPct, fill=factor(Year))) +
  geom_bar(stat="identity", position=position_dodge()) +
  geom_text(aes(label=n), vjust=-.5, color="black", size=3.5, position = position_dodge(0.9))+
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0:1), breaks = seq(0,1,.1)) +
  geom_hline(yintercept = .5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = overallwinrate, linetype = "dashed", color = "orange") +
  theme_bw() +
  ggtitle("Winrates by Month") +
  theme(plot.title = element_text(hjust = 0.5))
```
Historically, April and May have been good months fror my winrates. I also seem to play well later on in the year, though the number of games played are also on the lower end.
<br>
<br>

```{r}
#How Often Do I Get My Main Role
#2018
a1 <- length(which(dat18$Primary == dat18$Role))
b1 <- length(which(dat18$Secondary == dat18$Role))
a1 / nrow(dat18)
b1 / nrow(dat18)
```
For the most part, I get my Primary role.
<br>
<br>

```{r}
#What Is My Winrate By Role
#2018
#Primary
a2 <- which(dat18$Role == dat18$Primary)
mean(dat18[a2,]$Win1)

#Secondary
b2 <- which(dat18$Role == dat18$Secondary)
mean(dat18[b2,]$Win1)

#Jungle
jg <- which(dat18$Role == "JNG")
mean(dat18[jg,]$Win1)

#Support
sup <- which(dat18$Role == "SUP")
mean(dat18[sup,]$Win1)

#ADC
adc <- which(dat18$Role == "ADC")
mean(dat18[adc,]$Win1)
```
In 2018, my winrate on my Secondary role was an incredible 75%, however on a much smaller sample size. I also played better as a Jungler compared to Support. As ADC, I won every game but only played a few on that position.
<br>
<br>

```{r}
#2019
#Primary
a22 <- which(dat19$Role == dat19$Primary)
mean(dat19[a22,]$Win1)

#Secondary
b22 <- which(dat19$Role == dat19$Secondary)
mean(dat19[b22,]$Win1)

#Jungle
jg2 <- which(dat19$Role == "JNG")
mean(dat19[jg2,]$Win1)

#Support
sup2 <- which(dat19$Role == "SUP")
mean(dat19[sup2,]$Win1)

#ADC
adc2 <- which(dat19$Role == "ADC")
mean(dat19[adc2,]$Win1)

#MID
mid2 <- which(dat19$Role == "MID")
mean(dat19[mid2,]$Win1)
```
2019 was the opposite from 2018 as I won more games as Support and had a below 50% winrate as a Jungler.
<br>
<br>

```{r}
#2020
#Primary
a3 <- which(dat20$Role == dat20$Primary)
mean(dat20[a3,]$Win1)

#Secondary
b3 <- which(dat20$Role == dat20$Second)
mean(dat20[b3,]$Win1)

#Jungle
jg3 <- which(dat20$Role == "JNG")
mean(dat20[jg3,]$Win1)

#Support
sup3 <- which(dat20$Role == "SUP")
mean(dat20[sup3,]$Win1)
```
So far in 2020, my winrates across Primary and Secondary Role, Jungle and Support are all above the average of 50%.
<br>
<br>

```{r}
#Overall
dat18_2 <- dat18 %>% 
  select(Primary, Secondary, Role, Win1)
dat19_2 <- dat19 %>% 
  select(Primary, Secondary, Role, Win1)
dat20_2 <- dat20 %>% 
  select(Primary, Second, Role, Win1) %>% 
  rename(Secondary = Second)
tot <- rbind(dat18_2, dat19_2, dat20_2)

#Primary
a4 <- which(tot$Role == tot$Primary)
mean(tot[a4,]$Win1)

#Secondary
b4 <- which(tot$Role == tot$Secondary)
mean(tot[b4,]$Win1)

#Jungle
jg4 <- which(tot$Role == "JNG")
mean(tot[jg4,]$Win1)

#Support
sup4 <- which(tot$Role == "SUP")
mean(tot[sup4,]$Win1)

#Mid
mid4 <- which(tot$Role == "MID")
mean(tot[mid4,]$Win1)

#ADC
adc4 <- which(tot$Role == "ADC")
mean(tot[adc4,]$Win1)
```
Over the course of the last three years, Support has the highest winrate though the playcount is lower than Jungle.
<br>
<br>

## Champion Analysis
How well do I play on various champions?

```{r}
#Filter By Champion
#2018
mod18 <- dat18 %>% 
  group_by(Champion) %>% 
  summarise(KDA = round(((sum(Kills) + sum(Assists))/sum(Deaths)),3), Count = n(), WinPct = round(sum(Win1)/(sum(Win1) + sum(Loss1)),3)) %>% 
  arrange(desc(WinPct, KDA))
mod18

#2019
mod19 <- dat19 %>% 
  group_by(Champion) %>% 
  summarise(KDA = round(((sum(Kills) + sum(Assists))/sum(Deaths)),3), Count = n(), WinPct = round(sum(Win1)/(sum(Win1) + sum(Loss1)),3)) %>% 
  arrange(desc(WinPct, KDA))
mod19

#2020
mod20 <- dat20 %>% 
  group_by(Champion) %>% 
  summarise(KDA = round(((sum(Kills) + sum(Assists))/sum(Deaths)),3), Count = n(), WinPct = round(sum(Win1)/(sum(Win1) + sum(Loss1)),3)) %>% 
  arrange(desc(WinPct, KDA))
mod20
```


### Personal Performance by Champion
Sorting Champions by the number of times I recorded each grade.
```{r warning=FALSE, message=FALSE}
#2018
library(reshape2)
m18 <- dat18 %>% 
  select(Champion, Grade) %>% 
  group_by(Champion) %>% 
  mutate(Count = n()) %>% 
  arrange(desc(Count))

m18$Grade <- factor(m18$Grade,levels = c("S+","S","S-","A+","A","A-","B+","B","B-","C+","C","C-"))

dcast(m18, Champion~Grade, fun.aggregate = length)


#2019
m19 <- dat19 %>% 
  select(Champion, Grade) %>% 
  group_by(Champion) %>% 
  mutate(Count = n()) %>% 
  arrange(desc(Count))

m19$Grade <- factor(m19$Grade,levels = c("S+","S","S-","A+","A","A-","B+","B","B-","C+","C","C-"))

dcast(m19, Champion~Grade, fun.aggregate = length)


#2020
m20 <- dat20 %>% 
  select(Champion, Grade) %>% 
  group_by(Champion) %>% 
  mutate(Count = n()) %>% 
  arrange(desc(Count))

m20$Grade <- factor(m20$Grade,levels = c("S+","S","S-","A+","A","A-","B+","B","B-","C+","C","C-"))
m20$Champion <- factor(m20$Champion, levels = c("Zac","Elise",  "Trundle" ,  "Sejuani" ,  "Nautilus","Jarvan IV", "Rek'Sai" ))

dcast(m20, Champion~Grade, fun.aggregate = length)
```
<br>
<br>
Sorting Champions by the number of times I recorded each rank with 1 being the best.
```{r}
#Personal
#2019
mm19 <- dat19 %>% 
  select(Champion, `My Rank`) %>% 
  group_by(Champion) %>% 
  mutate(Count = n()) %>% 
  arrange(desc(Count))

dcast(mm19, Champion~`My Rank`, fun.aggregate = length)

mm19 %>% 
  group_by(Champion) %>% 
  summarise(AvgRank = mean(`My Rank`), Count = n()) %>% 
  arrange(AvgRank)


#2020
mm20 <- dat20 %>% 
  select(Champion, `My Rank`) %>% 
  group_by(Champion) %>% 
  mutate(Count = n()) %>% 
  arrange(desc(Count))

dcast(mm20, Champion~`My Rank`, fun.aggregate = length)

mm20 %>% 
  group_by(Champion) %>% 
  summarise(AvgRank = mean(`My Rank`), Count = n()) %>% 
  arrange(AvgRank)

dat20 %>% 
  select(Champion, `Team Rank`) %>%
  group_by(Champion) %>% 
  summarise(AvgTeamRank = mean(`Team Rank`), Count = n()) %>% 
  arrange(AvgTeamRank)
```


```{r}
#Plot KP vs KDA
ggplot(data = dat20, aes(x = KDA, y = `KP%`, color=factor(Win1))) +
  geom_point() + 
  geom_vline(xintercept = 3, linetype = "dashed", color = "red")
```
<br>
<br>
For the most part, if I have a KDA of 3 works as a simple cluster barrier to determine if I win the game or not.

<br>
<br>
```{r}
#Plot OP Score vs KDA
ggplot(data = dat20, aes(x = `OP Score`, y = `My Rank`, color=factor(Win1))) +
  geom_point() +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red")
```
<br>
<br>
Out of the 10 people in the game, if I rank #3 or higher, I will most likely win the game.